DSCI 445 - Algal Blooms

### 2020-12-14 paper.Rmd

---
title: "DSCI 445 - Algal Blooms"
author: "Dawson Eliasen, Boston Lee, Miles Austin"
date: "12/7/2020"
output:
  pdf_document: default
---

@@

# Background

@@

@@

The insitu data was not collected as frequently as the satellite imagery data.
To remedy this, an experimental approach was taken in which the FIXME data
was shifted by a certain number of days forward and backward. Intuitively, algae
blooms do not drastically alter the characteristics of a lake between consecutive days.
Interpolating surrounding observations with existing data allowed for many more
of the satellite observations to be used, even though there was no insitu data
accompanying the observations.

@@

@@

Comparing across different shifting thresholds, all three models included in
the experiment performed similarly.

@@

As we can see from Figure \ref{fig:thresh-performance}, the trend across merge
thresholds for each model is very similar. As can also be seen in Figure
\ref{fig:thresh-performance}, the optimal shifting threshold is three
days \ref{fig:thresh-performance}. It is also interesting to note that a shift of one day performed
worse than no shift, or a shift of two days. This is a counterintuitive result,
but it could be due to the tradeoff between the extra volume of data added by
larger shift thresholds versus the accuracy of the data at lower shifting
thresholds.

@@

All three models mentioned performed similarly, with no model greatly exceeding
the other two in terms of CV error. The best overall model, in terms of FIXME
error, was a random forest with a merge shift threshold of three days, that
included all features. The best model had a CV MSE of approximately 1.75, and a
test MSE of approximately 1.09.

>>>>

---
title: "DSCI 445 - Algal Blooms"
author: "Miles Austin, Dawson Eliasen, Boston Lee"
date: "12/14/2020"
output:
  pdf_document: default
---

@@

# Motivation

@@

![](images/map.jpeg)

![](images/HT_Climatology.jpeg)

![](images/ht_color_over_time.jpeg)

@@

The in situ data was not collected as frequently as the satellite imagery data.
To remedy this, an experimental approach was taken in which the FIXME data
was shifted by a certain number of days forward and backward. Intuitively, algae
blooms do not drastically alter the characteristics of a lake between consecutive days.
Interpolating surrounding observations with existing data allowed for many more
of the satellite observations to be used, even though there was no in situ data
accompanying the observations.

@@

Overall, random forest, gradient boost, and XGBoost performed similarly, with random forest
performing slightly better than the other two model types.

## Feature subsets and importance

Performance for all three models generally improved as more features were added.
The inclusion of the lake area feature resulted in the largest performance improvement.
\ref{fig:feature-importance} shows the feature importance statistics for a random forest
model fit to a dataset constructed with a merge threshold of 3 days and including all features.

```{python, echo=FALSE, fig.cap="Feature importance from random forest with all features included.\\label{fig:feature-importance}"}
plot_feature_importance()
```

@@

When comparing results across different merge threshold values, we see all three model
types perform similarly.

@@

As we can see from Figure \ref{fig:thresh-performance}, the trend across merge
thresholds for each model is very similar. Generally, performance decreases before increasing
and staying relatively high. It interesting to note that a shift of one day performed worse 
than no shift, or a shift of two days. This is a counter-intuitive result,
but it could be due to the tradeoff between the extra volume of data added by
larger shift thresholds versus the accuracy of the data at lower shifting
thresholds.

@@

All three models mentioned performed similarly, with no model greatly exceeding
the other two in terms of CV error. The best overall model, in terms of CV MSE
error, was a random forest with a merge shift threshold of three days, that
included all features. The best model had a CV MSE of approximately 1.75, and a
test set MSE of approximately 1.09.

## Predicted Values

Finally, we generate chlorophyll estimates for each of the available satellite images,
including satellite images of lakes for which there is no available in situ data.
\ref{fig:predicted-values} Shows the estimated chlorophyll value against the dominant
wavelength of the satellite image. We can see that the highest chlorophyll estimates
are attributed to satellite images which have a green dominant wavelength. However, we
are generating predictions of chlorophyll values exceeding $70 \frac{\mu g}{mL}$, which
is most likely unreasonable. This is likely an artifact of the extreme observations included
in the training data with chlorophyll measurements exceeding $300 \frac{\mu g}{mL}$.

```{python, echo=FALSE, fig.cap="Predicted values vs. dominant wavelength.\\label{fig:predicted-values}"}
plot_predicted_values()
```

## Conclusion

We have described a process for using non-parametric machine learning methods to estimate
chlorophyll concentrations in lakes from satellite images. This process broadly consists of
these steps.

1. Collect satellite image data.
2. Process satellite image data to create features.
3. Associate image features with in situ measurements, allowing some small number of days in error.
4. $\log$-transform the chlorophyll measurements to unskew the distribution.
5. Fit a model to these data, with $\log$-transform chlorophyll measurements as the response variable and satellite features as the predictor variables.
6. Use this model to generate predicted $\log (\text{chlorophyll})$ values for each satellite image.
7. Exponentiate the predicted values to return them to original scale.

Using this process, we investigated the effectiveness of different model types, feature subsets,
and merge threshold values and found the best results with a random forest model using all features
and a merge threshold value equal to three days. This model had a 5-fold cross validation MSE of 
1.75, and the MSE on the holdout test set was 1.05. Based on these error measurements,
we have demonstrated that these models are capable of estimating chlorophyll concentrations
with practical accuracy. This level of accuracy is likely at least partially attributable to the fact
that the training data includes repeated sampling of a handful of lakes, and as a result the error
measurements are measuring the model's ability to estimate chlorophyll concentrations from *unseen*
images of *familiar* lakes. To evaluate the predictive accuracy of these models on *unfamiliar* lakes,
an analyst would require a dataset observing a much larger number and range of lakes. As mentioned
previously, only measurements from municipal water sources were available.

In future exploration, we hope to address the extreme predicted values by removing observations with a
chlorophyll concentration exceeding $15 \frac{\mu g}{mL}$. We expect this change to result in a much
more reasonable range of predicted values. We also think that, based on the importance of the lake area
feature, the inclusion of a feature describing lake elevation will be important
